---
title: "OPA Data Exploratory Analysis"
author: "Team 6"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Data Sources

- **Main table**: `musa5090s25-team6.core.opa_properties` (via BigQuery)
- **Supplementary table**: `musa5090s25-team6.core.opa_assessments`
- **Original dataset**: [Philadelphia Properties and Assessment History - OpenDataPhilly](https://opendataphilly.org/dataset/opa-property-assessments)

---

## 1. Load Required Packages

```{r libraries}
library(tidyverse)
library(skimr)
library(corrplot)
library(ggplot2)
library(lubridate)
library(scales)
library(gridExtra)
```

---

## 2. Connect to BigQuery and Load Data

```{r load-data}
# Source: musa5090s26-team6.core.opa_properties
df_raw <- read.csv("opa_properties.csv")

cat("Raw row count:", nrow(df_raw), "\n")
cat("Number of columns:", ncol(df_raw), "\n")
```

---

## 3. Data Cleaning

### 3.1 Filter Out Anomalously Low Sale Prices

Properties with very low sale prices (e.g. $1) are likely family transfers or administrative transactions and are not representative of true market value.

```{r filter-low-price}
# Check the price distribution before filtering
summary(df_raw$sale_price)

# Remove records where sale_price <= 100
df_clean <- df_raw |>
  filter(
    sale_price > 100,
    sale_price < 2000000
  )

cat("Rows after filtering low prices:", nrow(df_clean), "\n")
cat("Records removed:", nrow(df_raw) - nrow(df_clean), "\n")
```

### 3.2 Filter Out Bundle Sales

Properties sold on the exact same day at the exact same price are likely part of a bundle sale. In those cases, `sale_price` represents the price of the entire bundle, not any individual property.

```{r filter-bundle}
# Identify bundle sales: same sale_date AND same sale_price
df_clean <- df_clean |>
  group_by(sale_date, sale_price) |>
  mutate(bundle_count = n()) |>
  ungroup() |>
  filter(bundle_count == 1) |>
  select(-bundle_count)

cat("Rows after filtering bundle sales:", nrow(df_clean), "\n")
```

### 3.3 Type Conversion and Derived Features

```{r type-conversion}
df_clean <- df_clean |>
  mutate(
    # --- Date features ---
    sale_date = as.Date(sale_date),
    sale_year = year(sale_date),
    days_since_sale = as.numeric(Sys.Date() - sale_date),

    # --- Derived numeric features ---
    property_age = 2025 - year_built,
    log_sale_price = log(sale_price),
    log_livable_area = log(total_livable_area + 1),

    # --- Fix anomalous property age ---
    property_age = ifelse(property_age < 0 | property_age > 200, NA, property_age),

    # --- Condition fields to numeric ---
    exterior_condition = as.numeric(exterior_condition),
    interior_condition = as.numeric(interior_condition),

    # --- Simplify zoning into broad categories ---
    zoning_group = case_when(
      str_starts(zoning, "RM") ~ "Residential_Multi",
      str_starts(zoning, "RSA") | str_starts(zoning, "RSD") ~ "Residential_Single",
      str_starts(zoning, "CMX") | str_starts(zoning, "CA") ~ "Commercial",
      str_starts(zoning, "I") ~ "Industrial",
      TRUE ~ "Other"
    ),

    # --- zip_code as character for encoding ---
    zip_code = as.character(zip_code)
  ) |>
  # --- Remove remaining anomalous values ---
  filter(
    sale_price > 100,
    sale_price < 2000000,
    total_livable_area > 0 | is.na(total_livable_area)
  )
```

---

## 4. Exploratory Data Analysis

### 4.1 Target Variable Distribution: sale_price

```{r sale-price-dist, fig.width=10, fig.height=5}
p1 <- ggplot(df_clean, aes(x = sale_price)) +
  geom_histogram(bins = 100, fill = "#4A90D9", color = "white", alpha = 0.8) +
  scale_x_continuous(labels = dollar_format()) +
  labs(title = "Sale Price Distribution (Raw)",
       x = "Sale Price", y = "Count") +
  theme_minimal()

p2 <- ggplot(df_clean, aes(x = log_sale_price)) +
  geom_histogram(bins = 80, fill = "#E8593C", color = "white", alpha = 0.8) +
  labs(title = "Log(Sale Price) Distribution",
       x = "Log(Sale Price)", y = "Count") +
  theme_minimal()

grid.arrange(p1, p2, ncol = 2)
```

### 4.2 Time Trend: Sale Year vs Median Sale Price

More recent sales are likely to carry a stronger signal for current property values.

```{r time-trend, fig.width=10, fig.height=5}
df_clean |>
  filter(sale_year >= 2010) |>
  group_by(sale_year) |>
  summarise(median_price = median(sale_price, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = sale_year, y = median_price)) +
  geom_line(color = "#4A90D9", size = 1.2) +
  geom_point(color = "#E8593C", size = 3) +
  scale_y_continuous(labels = dollar_format()) +
  labs(title = "Median Sale Price by Year",
       x = "Year", y = "Median Sale Price") +
  theme_minimal()
```

### 4.3 Total Livable Area vs Sale Price

```{r area-price, fig.width=8, fig.height=6}
df_clean |>
  filter(total_livable_area > 0, total_livable_area < 5000) |>
  sample_n(min(5000, nrow(df_clean))) |>
  ggplot(aes(x = total_livable_area, y = sale_price)) +
  geom_point(alpha = 0.3, color = "#4A90D9", size = 0.8) +
  geom_smooth(method = "lm", color = "#E8593C", se = TRUE) +
  scale_y_continuous(labels = dollar_format()) +
  labs(title = "Total Livable Area vs Sale Price",
       x = "Total Livable Area (sqft)", y = "Sale Price") +
  theme_minimal()
```

### 4.4 Property Age vs Sale Price

```{r age-price, fig.width=8, fig.height=6}
df_clean |>
  filter(!is.na(property_age), property_age >= 0, property_age < 200) |>
  ggplot(aes(x = property_age, y = log_sale_price)) +
  geom_point(alpha = 0.2, color = "#4A90D9", size = 0.8) +
  geom_smooth(method = "loess", color = "#E8593C", se = TRUE) +
  labs(title = "Property Age vs Log(Sale Price)",
       x = "Property Age (years)", y = "Log(Sale Price)") +
  theme_minimal()
```

### 4.5 Median Sale Price by Zip Code (Top 20)

```{r zip-price, fig.width=12, fig.height=6}
df_clean |>
  filter(!is.na(zip_code)) |>
  group_by(zip_code) |>
  filter(n() >= 30) |>
  summarise(median_price = median(sale_price, na.rm = TRUE),
            count = n(), .groups = "drop") |>
  arrange(desc(median_price)) |>
  slice_head(n = 20) |>
  ggplot(aes(x = reorder(zip_code, median_price), y = median_price)) +
  geom_col(fill = "#4A90D9", alpha = 0.8) +
  scale_y_continuous(labels = dollar_format()) +
  coord_flip() +
  labs(title = "Top 20 Zip Codes by Median Sale Price",
       x = "Zip Code", y = "Median Sale Price") +
  theme_minimal()
```

### 4.6 Feature Correlation Matrix

```{r correlation, fig.width=10, fig.height=8}
numeric_cols <- df_clean |>
  select(
    sale_price, log_sale_price,
    total_livable_area, total_area,
    number_of_bedrooms, number_of_bathrooms,
    property_age, days_since_sale,
    exterior_condition, interior_condition,
    garage_spaces, fireplaces
  ) |>
  drop_na()

cor_matrix <- cor(numeric_cols, use = "complete.obs")
corrplot(cor_matrix, method = "color", type = "upper",
         tl.cex = 0.8, addCoef.col = "black", number.cex = 0.7,
         col = colorRampPalette(c("#E8593C", "white", "#4A90D9"))(200),
         title = "Feature Correlation Matrix", mar = c(0,0,2,0))
```

### 4.7 Missing Value Analysis

```{r missing-values, fig.width=10, fig.height=6}
missing_summary <- df_clean |>
  summarise(across(everything(), ~sum(is.na(.))/n()*100)) |>
  pivot_longer(everything(), names_to = "variable", values_to = "missing_pct") |>
  filter(missing_pct > 0) |>
  arrange(desc(missing_pct))

ggplot(missing_summary, aes(x = reorder(variable, missing_pct), y = missing_pct)) +
  geom_col(fill = "#E8593C", alpha = 0.8) +
  coord_flip() +
  labs(title = "Missing Value Rate by Feature (%)",
       x = "Feature", y = "Missing (%)") +
  theme_minimal()
```

---

## 5. Summary Statistics

```{r skim}
skim(df_clean |> select(
  sale_price, log_sale_price, total_livable_area,
  number_of_bedrooms, number_of_bathrooms,
  property_age, days_since_sale,
  exterior_condition, interior_condition
))
```

---

## 6. Key Findings

1. **Outlier filtering is critical**: Raw data contained sale prices 
   up to $968M. Filtering to $100 < sale_price < $2M removed ~36% 
   of records but dramatically improved feature correlations — for 
   example, the correlation between total_livable_area and sale_price 
   improved from 0.05 to 0.44 after filtering.

2. **Recency is the strongest predictor**: days_since_sale has the 
   highest correlation with log_sale_price (-0.56). Philadelphia 
   median sale prices rose sharply from ~$120K in 2010 to ~$260K 
   in 2021, and have since stabilized. More recent sales carry 
   stronger signals for current property values.

3. **Condition outperforms size**: interior_condition (-0.41) and 
   exterior_condition (-0.33) are stronger predictors than 
   total_livable_area (0.31) when predicting log_sale_price.

4. **Location drives large price variation**: Zip code 19118 has a 
   median sale price of ~$450K while 19154 is ~$175K — a 2.5x 
   difference purely from location. zip_code must be included 
   as a feature, with target encoding recommended for modeling.

5. **number_of_bedrooms is a weak predictor** (correlation: 0.09) 
   despite being intuitively important. This is likely because 
   bedrooms are highly correlated with bathrooms (r=0.64), and 
   bathrooms independently carry more predictive power.

6. **High missing rates for amenity features**: garage_spaces, 
   fireplaces, number_of_bathrooms, and number_of_bedrooms all 
   have ~10% missing values. basements and central_air are 
   excluded entirely due to >40% missing rates.

**Recommended features for the assessment model (Issue #11):**
- Time: days_since_sale
- Size: log_livable_area, total_area
- Condition: exterior_condition, interior_condition, property_age
- Location: zip_code (target encoded)
- Amenities: number_of_bathrooms, number_of_bedrooms, 
  garage_spaces, fireplaces

---

*All analysis artifacts are saved in the `eda/` folder. Data is sourced from BigQuery and raw data files are not committed to this repository.*
